### 2017-06-30 draft/unique-linear.tex
effect propagation

\begin{mathpar}
  \inferrule{
    A = A_1 \OVERRIDE \dots \OVERRIDE A_n \\
    A_i \vdash e_i : t_i 
  }{
    A \vdash \REC{ \overline{a_i=e_i}} : \REC{ \overline{a_i :^{P_i} t_i}}
  }

@@

  \inferrule{
    A_0, x:t \vdash e : t' \\
    A = A_0^*
  }{
    A \vdash \LAM x t e : (t \to t')
  }
\end{mathpar}
Function closures have ``infinite'' splitting $(a :^P t)^*$ into $(a :^Q t)$.
\begin{itemize}
\item $P \le \{\READ\}$ propagate $Q = \READ$, no action required
\item $\WRITE \in P$: propagate $Q = \READ$, insert copy into closure
\end{itemize}

Function application has no particular restrictions, it seems.
\begin{mathpar}
  \inferrule{
    A_0 \vdash e_0 : (t_1 \to t_2) \\
    A_1 \vdash e_1 : t_1 \\
    A = A_0 \OVERRIDE A_1
  }{
    A \vdash \APP{e_0}e_1 : t_2
  }
\end{mathpar}

Look at combining $a :^O t \OVERRIDE a:^P t$ upwards to $a : ^Q t$
\begin{itemize}
\item if $O=\emptyset$ or $P=\emptyset$ (field only used in one branch), then $Q = O \cup P$, no
  action required
\item if $O\subseteq \{\READ\}$: $Q = O \cup P$,
  no further action required as potential writing in $P$ happens after reading
\item if $\WRITE \in O$, $P = \{\READ\}$: two choices, not clear
  which is better
  \begin{itemize}
  \item $Q = \READ$ insert copy, propagate left to the write operation
  \item $Q = \WRITE$ insert copy, propagate right to the read operation
  \end{itemize}
\item if $\WRITE\in O$, $\WRITE \in P$: $Q = \WRITE$ insert copy,
  propagate left (or right)
\end{itemize}

\begin{mathpar}
  \inferrule
  { A_1 \vdash e_1 : t_1 \\
    A_2, x : t_1 \vdash e_2 : t_2 \\
    A = A_1 \OVERRIDE A_2
  }
  { A \vdash \LET x{e_1} e_2 : t}
\end{mathpar}

>>>>

Effects attached to single record labels, intuitively $P, Q \subseteq
\{ \READ, \WRITE \}$ but it does not make sense to distinguish
$\{\WRITE, \READ\}$ from $\{\WRITE\}$, so we take the three element lattice
\begin{align*}
  O, P, Q &::= \NOTHING \mid \READ \mid \WRITE
  & \text{ordered by } \NOTHING
  &\sqsubset \READ \sqsubset \WRITE
\end{align*}

\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{
      A = A_1 \OVERRIDE \dots \OVERRIDE A_n \\
      A_i \vdash e_i : t_i }{ A \vdash \REC{ \overline{a_i=e_i}} :
      \REC{ \overline{a_i :^{P_i} t_i}} }

@@

    \inferrule{
      A_0, x:t \vdash e : t' \\
      A = A_0^* }{ A \vdash \LAM x t e : (t \to t') }

  \inferrule{
    A_0 \vdash e_0 : (t_1 \to t_2) \\
    A_1 \vdash e_1 : t_1 \\
    A = A_0 \OVERRIDE A_1
  }{
    A \vdash \APP{e_0}e_1 : t_2
  }

  \inferrule
  { A_1 \vdash e_1 : t_1 \\
    A_2, x : t_1 \vdash e_2 : t_2 \\
    A = A_1 \OVERRIDE A_2
  }
  { A \vdash \LET x{e_1} e_2 : t}
  \end{mathpar}
  \caption{Propagation rules}
  \label{fig:field-effects}
\end{figure}
Function closures have ``infinite'' splitting $(a :^P t)^*$ into $(a :^Q t)$.
\begin{itemize}
\item $P = \NOTHING$ propagate $Q=\NOTHING$.
\item $P = \READ$ propagate $Q = \WRITE$, no action required inside closure.
\item $P = \WRITE$ propagate $Q = \WRITE$, perform copy inside of closure.
\end{itemize}
 Propagating $\READ$ is not safe in both cases because that enables a write operation on the same
 record after the closure has been constructed. Alternatively, we could have another READONLY mode,
 which prevents subsequent write operations.

Function application has no particular restrictions, it seems.
Let improves on lambda combined with app because infinite splitting is avoided.

Look at combining $a :^O t \OVERRIDE a:^P t$ upwards to $a : ^Q t$:

$Q= O \OVERRIDE P$ is defined by:
\begin{itemize}
\item if $O=\NOTHING$ or $P=\NOTHING$ (field only used in one branch), then $Q = O \sqcup P$, no
  action required
\item if $O\sqsubseteq \READ$: $Q = O \sqcup P$,
  no further action required as potential writing in $P$ happens after reading
\item if $O = \WRITE$, $P = \READ$, then there are two choices
  \begin{itemize}
  \item $Q = \READ$ insert copy, propagate left to the write operation
  \item $Q = \WRITE$ insert copy, propagate right to the read operation
  \end{itemize}
  First option is preferred because it keeps writes local and prevents
  further copies downstream.
\item if $O=\WRITE$, $P =\WRITE$: $Q = \WRITE$ insert copy,
  propagate left (or right)
\end{itemize}
We need an operation that returns $\READ$ if a copy is required and
$\NOTHING$ otherwise.
\begin{align*}
  \COPY (O, P) &=
                 \begin{cases}
                   \READ & O=\WRITE \wedge \READ \sqsubseteq P \\
                   \NOTHING & \text{otherwise}
                 \end{cases}
  \\
  \COPY (\overline{O}, \overline{P})
               &= \bigsqcup_i \COPY (O_i, P_i)
  \\
  \COPY (\NUM, \NUM)
               & = \NOTHING
  \\
  \COPY (t_1 \to t_2, t_1 \to t_2)
               & = \NOTHING
  \\
  \COPY (\REC{ \overline{a_i :^{O_i} t_i'}}, \REC{ \overline{a_i :^{P_i} t_i''}})
  &=\COPY (\overline{O_i}, \overline{P_i})
    \sqcup \bigsqcup_i \overline{\COPY (t_i', t_i'')}
\end{align*}

Combining environments: pointwise

Combining types and records
\begin{mathpar}
  \inferrule{}{ \NUM = \NUM \OVERRIDE \NUM}

  \inferrule{}{t_1 \to t_2 = t_1 \to t_2 \OVERRIDE t_1 \to t_2}
  \\
  \inferrule{
    Q_i = (O_i \OVERRIDE P_i)
    \sqcup \COPY (\overline{O_i}, \overline{P_i})
    \sqcup \bigsqcup_i \overline{\COPY (t_i', t_i'')} \\
    t_i = t_i' \OVERRIDE t_i''
  }{
    \REC{ \overline{a_i :^{Q_i} t_i}}
    =
    \REC{ \overline{a_i :^{O_i} t_i'}}
    \OVERRIDE
    \REC{ \overline{a_i :^{P_i} t_i''}}
  }
\end{mathpar}

\begin{mathpar}
  \inferrule
  {}
  {
    \REC{ {a :^{Q_a} t_a}, {b :^{Q_b} t_b}}
    =
    \REC{ {a :^{O_a} t_a}, {b :^{O_b} t_b}}
    \OVERRIDE
    \REC{ {a :^{P_a} t_a}, {b :^{P_b} t_b}}
  }
\end{mathpar}

\clearpage
\begin{mathpar}
  \begin{array}{|c|c||c|c||c|c|l}
    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
    \hline
    P_a &\NOTHING& \NOTHING & \NOTHING & P_a & \NOTHING & \\
    O_a & \NOTHING& O_a & \NOTHING & \NOTHING & \NOTHING & \\
    P_a\sqcup\READ   &\NOTHING& \READ & \NOTHING & P_a & \NOTHING & \\
    \READ&\READ& \WRITE & \NOTHING & \READ & \NOTHING
                       &  \text{copy, propagate left to write}\\
    \WRITE &\READ& \WRITE & \NOTHING & \WRITE & \NOTHING
                       &\text{copy, propagate left to write} \\
  \end{array}

  \begin{array}{|c|c||c|c||c|c|l}
    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
    \hline
    P_a & P_b & \NOTHING & \NOTHING & P_a & P_b & \\
    O_a & P_b & O_a & \NOTHING & \NOTHING & P_b & \\
    P_a\sqcup\READ   &P_b& \READ & \NOTHING & P_a & P_b & \\
    \READ&P_b\sqcup\READ& \WRITE & \NOTHING & \READ & P_b
                       &  \text{copy, propagate left to write}\\
    \WRITE & P_b\sqcup\READ & \WRITE & \NOTHING & \WRITE & P_b
                       &\text{copy, propagate left to write} \\
  \end{array}

  \begin{array}{|c|c||c|c||c|c|l}
    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
    \hline
    P_a & O_b & \NOTHING & O_b & P_a & \NOTHING & \\
    O_a & O_b & O_a & O_b & \NOTHING & \NOTHING & \\
    P_a\sqcup\READ  & O_b & \READ & O_b & P_a & \NOTHING & \\
    \READ&O_b \sqcup\READ& \WRITE & O_b & \READ & \NOTHING
                       &  \text{copy, propagate left to write}\\
    \WRITE &O_b\sqcup\READ& \WRITE & O_b & \WRITE & \NOTHING
                           &\text{copy, propagate left to write} \\
  \end{array}

\begin{array}{|c|c||c|c||c|c|l}
    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
    \hline
    P_a &P_b\sqcup\READ& \NOTHING & \READ & P_a & P_b & \\
    O_a & P_b\sqcup\READ& O_a & \READ & \NOTHING & P_b & \\
    P_a\sqcup\READ   &P_b\sqcup\READ& \READ & \READ & P_a & P_b & \\
    \READ&P_b\sqcup\READ& \WRITE & \READ & \READ & P_b
                       &  \text{copy, propagate left to write}\\
    \WRITE &P_b\sqcup\READ& \WRITE & \READ & \WRITE & P_b
                       &\text{copy, propagate left to write} \\
  \end{array}

  \begin{array}{|c|c||c|c||c|c|l}
    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
    \hline
    P_a\sqcup\READ &\READ& \NOTHING & \WRITE & P_a & \READ
                           & \text{copy, propagate left to write} \\
    \READ & \READ& O_a & \WRITE & \NOTHING & \READ
                           & \text{copy, propagate left to write} \\
    P_a\sqcup\READ   &\READ& \READ & \WRITE & P_a & \READ
                           &  \text{copy, propagate left to write}\\
    \READ&\READ& \WRITE & \WRITE & \READ & \READ
                       &  \text{copy, propagate left to write}\\
    \WRITE &\READ& \WRITE & \WRITE & \WRITE & \READ
                       &\text{copy, propagate left to write} \\
  \end{array}

  \begin{array}{|c|c||c|c||c|c|l}
    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
    \hline
    P_a\sqcup\READ &\WRITE& \NOTHING & \WRITE & P_a & \WRITE
                           & \text{copy, propagate left to write} \\
    \READ & \WRITE& O_a & \WRITE & \NOTHING & \WRITE
                           & \text{copy, propagate left to write} \\
    P_a\sqcup\READ   &\WRITE& \READ & \WRITE & P_a & \WRITE
                           &  \text{copy, propagate left to write}\\
    \READ&\WRITE& \WRITE & \WRITE & \READ & \WRITE
                       &  \text{copy, propagate left to write}\\
    \WRITE &\WRITE& \WRITE & \WRITE & \WRITE & \WRITE
                       &\text{copy, propagate left to write} \\
  \end{array}
\end{mathpar}
A copy operation reads \textbf{every} field. Hence, a copy operation
caused by one field implies a read effect on all fields.
